---
title: "Divorce Prediction with Machine Learning"
output: html_document
---

```{r setup, include=FALSE}
library(caret)
library(ranger)
library(ggplot2)
library(corrplot)
```

## Data

The project uses the UCI Divorce Predictors dataset. Place the semicolon-delimited file at `data/divorce.csv`.

```{r}
set.seed(1821)

data_path <- file.path("data", "divorce.csv")
if (!file.exists(data_path)) {
  stop("Missing data/divorce.csv. See data/README.md for the expected file.")
}

divorce <- read.csv(data_path, sep = ";", check.names = FALSE)
if (!"Class" %in% names(divorce)) {
  stop("Expected a target column named 'Class'.")
}

divorce$Class <- factor(divorce$Class)
if (nlevels(divorce$Class) != 2) {
  stop("This workflow expects a binary Class target.")
}

str(divorce)
table(divorce$Class)
```

## Stratified hold-out split

The small dataset makes evaluation especially sensitive to a single split, so the target is stratified and the hold-out set is kept untouched until final evaluation.

```{r}
set.seed(1821)
train_index <- createDataPartition(divorce$Class, p = 0.75, list = FALSE)
train_data <- divorce[train_index, ]
test_data <- divorce[-train_index, ]
```

## Training-only exploratory structure

Correlation and PCA are calculated from the training partition only. The test set does not influence these exploratory transformations.

```{r, fig.width=9, fig.height=8}
predictor_names <- setdiff(names(train_data), "Class")
train_correlation <- cor(train_data[predictor_names], method = "spearman")
corrplot(train_correlation, type = "lower", tl.cex = 0.45)
```

```{r}
pca_model <- prcomp(
  train_data[predictor_names],
  center = TRUE,
  scale. = TRUE
)

train_pca <- as.data.frame(predict(pca_model, newdata = train_data[predictor_names]))
train_pca$Class <- train_data$Class

test_pca <- as.data.frame(predict(pca_model, newdata = test_data[predictor_names]))
test_pca$Class <- test_data$Class

summary(pca_model)
```

```{r}
ggplot(train_pca, aes(PC1, PC2, color = Class)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "Training-set PCA projection")
```

## Model training

Two interpretable baselines are compared using repeated cross-validation **inside the training partition**. The held-out test set is not used for model selection.

```{r}
control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 5,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(1821)
logistic_model <- train(
  Class ~ .,
  data = train_data,
  method = "glm",
  family = binomial(),
  trControl = control,
  metric = "Accuracy"
)

set.seed(1821)
rf_model <- train(
  Class ~ .,
  data = train_data,
  method = "ranger",
  trControl = control,
  tuneLength = 5,
  metric = "Accuracy",
  num.trees = 500,
  importance = "permutation"
)

logistic_model
rf_model
```

## Held-out evaluation

```{r}
logistic_predictions <- predict(logistic_model, newdata = test_data)
rf_predictions <- predict(rf_model, newdata = test_data)

logistic_confusion <- confusionMatrix(logistic_predictions, test_data$Class)
rf_confusion <- confusionMatrix(rf_predictions, test_data$Class)

logistic_confusion
rf_confusion
```

## Variable importance

```{r}
plot(varImp(rf_model), top = 20, main = "Random Forest predictor importance")
```

## Interpretation note

This dataset contains only about 170 observations. Very high accuracy on one split should therefore not be interpreted as evidence that the model will generalize equally well to new populations. The held-out result is reported as an educational demonstration rather than a clinical, legal or relationship-assessment tool.

## Session information

```{r}
sessionInfo()
```
